---
title: "Predicting Happiness Scores"
by: "Oana Doherty"
output: pdf_document
date: "2022-10-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

This project will analyse the data to seek possible correlations between happiness scores and other criteria being test.  I will look at all countries test, and then consider separately the happiest and least happy countries.  Finally, I will try to predict the happiness score of 175 countries.  I will do this using a variety of methods: by guessing, by using a correlation line, and finally by using the Root Mean Squared Error, RMSE.  

The data scores 12 categories, one of which is the overall happiness score, over the course of seven years, from 2015 through 2022. The raw data is located at https://www.kaggle.com/datasets/mathurinache/world-happiness-report.  I combined all years into one file which is downloaded and read by the script.  It can also be viewed at https://github.com/Oanalkd/Happiness/blob/main/Happiness.csv.

What we will see is that on average happiness scores are increasing, although for some of the unhappiest ones, they are not.  On average, not at the highest and lowest scores, the highest correlation occur between happiness and GDP, health, and freedom.  A negative correlation occurs between happiness and generosity at both the top and bottom of the spectrum.  

To predict the happiness levels, I train the model on years 2015 - 2021 and test it on year 2022 for an rmse of .41.   


```{r}
#install required packages  
if(!require(data.table)) install.packages("tinytex", repos = "http://cran.us.r-project.org")
tinytex::install_tinytex(force = TRUE)
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(gridExtra)) install.packages("gridExtra", repos = "http://cran.us.r-project.org")


# remove all variables
rm (list=ls())

#Reading the data shows that there are 1230 records and 13 dimensions. 
url<-"https://raw.githubusercontent.com/Oanalkd/Happiness/main/Happiness.csv"
Happiness<-read.csv(url)
dim(Happiness)
names(Happiness)

#Create new fields that will be needed
Happiness$CtryYear<-paste(Happiness$Year, Happiness$Country, sep="_")
Happiness$GdpHealth<-(Happiness$Economy..GDP.per.Capita.+Happiness$Health..Life.Expectancy.)/2
Happiness$GdpHealthFree<-(Happiness$Economy..GDP.per.Capita.+Happiness$Health..Life.Expectancy.+Happiness$Freedom)/3


#There are 175 distinct countries
Happiness%>%summarise(countries=n_distinct(Country))

#Which country/year have the highest/lower scores
Happiness$CtryYear[which.max(Happiness$Happiness.Score)]
Happiness$CtryYear[which.min(Happiness$Happiness.Score)]

#Distribution of happiness rating by country AND year
plot1<-Happiness %>% group_by(Country)%>%
  ggplot(aes(Happiness.Score)) + 
  geom_histogram(fill = "light blue")+
  ggtitle("Score by Year and Country")
plot1


#We can see that the distribution is close to normal and therefore can be well described by the mean and standard deviation. 
plot2<-ggplot(data=Happiness,aes(sample=Happiness.Score))+stat_qq()+ggtitle("QQ PLot - Test for Normality")
plot2


#Obtain the mean Happiness score by Country only
MeanHappy<-Happiness%>% group_by(Country)%>% summarise(mean(Happiness.Score))
#Rename columns
colnames(MeanHappy)<-c("Country","Score")
plot3<-MeanHappy %>%ggplot(aes(Score)) +   geom_histogram(fill = "light blue")+  ggtitle("Score by Country")
grid.arrange(plot1,plot2,plot3,nrow=3)

#Top most and bottom most  scores
MeanHappy%>%top_n(5, Score)
MeanHappy%>%top_n(-5, Score)


#Obtain mean and sd of happiness score
summary <- Happiness %>%summarize(mean = mean(Happiness$Happiness.Score), sd = sd(Happiness$Happiness.Score))
summary

#Now we look at the overall happiness levels across years.  They are increasing.
MeanHappyYear<-Happiness%>% group_by(Year)%>% summarise(mean(Happiness.Score))
#Rename columns
colnames(MeanHappyYear)<-c("Year","Score")
MeanHappyYear
plot1<-MeanHappyYear%>%ggplot(aes(Year, Score)) + geom_line()
plot1


#We want to see which countries show up in the top 10 for each year.  
top<-MeanHappy%>%arrange(desc(Score))%>%slice(1:10)
top_2015<-Happiness%>%filter(Year==2015)%>%arrange(desc(Happiness.Score))%>%slice(1:10)
top_2016<-Happiness%>%filter(Year==2016)%>%arrange(desc(Happiness.Score))%>%slice(1:10)
top_2017<-Happiness%>%filter(Year==2017)%>%arrange(desc(Happiness.Score))%>%slice(1:10)
top_2018<-Happiness%>%filter(Year==2018)%>%arrange(desc(Happiness.Score))%>%slice(1:10)
top_2019<-Happiness%>%filter(Year==2019)%>%arrange(desc(Happiness.Score))%>%slice(1:10)
top_2020<-Happiness%>%filter(Year==2020)%>%arrange(desc(Happiness.Score))%>%slice(1:10)
top_2021<-Happiness%>%filter(Year==2021)%>%arrange(desc(Happiness.Score))%>%slice(1:10)
top_2022<-Happiness%>%filter(Year==2022)%>%arrange(desc(Happiness.Score))%>%slice(1:10)
top_allyears<-rbind(top_2015,top_2016,top_2017,top_2018,top_2019,top_2020,top_2021,top_2022)
#Countries which show up in the top 10 happiest countries each year, will show up 8 times in this table
table(top_allyears$Country)

#We will select countries which were in top 8 across all years
top_8_all_years<-top_allyears%>%filter(Country %in% c("Denmark","Finland","Iceland","Netherlands","New Zealand","Norway","Sweden","Switzerland"))
plot1<-top_8_all_years %>%filter(Year==2015)%>%ggplot(aes(x=Country,y=Happiness.Score))+geom_point() + ggtitle("Top 8 Happiest - 2015 ")+geom_text(aes(x=Country,y=Happiness.Score,label=Country))+theme(axis.text.x=element_blank(),axis.ticks.x=element_blank() )+labs(x=NULL,y=NULL)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())
plot2<-top_8_all_years %>%filter(Year==2016)%>%ggplot(aes(x=Country,y=Happiness.Score))+geom_point() + ggtitle("Top 8 Happiest - 2016 ")+geom_text(aes(x=Country,y=Happiness.Score,label=Country))+theme(axis.text.x=element_blank(),axis.ticks.x=element_blank() )+labs(x=NULL,y=NULL)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())
plot3<-top_8_all_years %>%filter(Year==2017)%>%ggplot(aes(x=Country,y=Happiness.Score))+geom_point() + ggtitle("Top 8 Happiest - 2017 ")+geom_text(aes(x=Country,y=Happiness.Score,label=Country))+theme(axis.text.x=element_blank(),axis.ticks.x=element_blank() )+labs(x=NULL,y=NULL)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())
plot4<-top_8_all_years %>%filter(Year==2018)%>%ggplot(aes(x=Country,y=Happiness.Score))+geom_point() + ggtitle("Top 8 Happiest - 2018 ")+geom_text(aes(x=Country,y=Happiness.Score,label=Country))+theme(axis.text.x=element_blank(),axis.ticks.x=element_blank() )+labs(x=NULL,y=NULL)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())
plot5<-top_8_all_years %>%filter(Year==2019)%>%ggplot(aes(x=Country,y=Happiness.Score))+geom_point() + ggtitle("Top 8 Happiest - 2019 ")+geom_text(aes(x=Country,y=Happiness.Score,label=Country))+theme(axis.text.x=element_blank(),axis.ticks.x=element_blank() )+labs(x=NULL,y=NULL)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())
plot6<-top_8_all_years %>%filter(Year==2020)%>%ggplot(aes(x=Country,y=Happiness.Score))+geom_point() + ggtitle("Top 8 Happiest - 2020 ")+geom_text(aes(x=Country,y=Happiness.Score,label=Country))+theme(axis.text.x=element_blank(),axis.ticks.x=element_blank() )+labs(x=NULL,y=NULL)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())
plot7<-top_8_all_years %>%filter(Year==2021)%>%ggplot(aes(x=Country,y=Happiness.Score))+geom_point() + ggtitle("Top 8 Happiest - 2021 ")+geom_text(aes(x=Country,y=Happiness.Score,label=Country))+theme(axis.text.x=element_blank(),axis.ticks.x=element_blank() )+labs(x=NULL,y=NULL)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())
plot8<-top_8_all_years %>%filter(Year==2022)%>%ggplot(aes(x=Country,y=Happiness.Score))+geom_point() + ggtitle("Top 8 Happiest - 2022 ")+geom_text(aes(x=Country,y=Happiness.Score,label=Country))+theme(axis.text.x=element_blank(),axis.ticks.x=element_blank() )+labs(x=NULL,y=NULL)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())
# we can see the same countries in top 8 happiest in each year
grid.arrange(plot1,plot2,plot3,plot4,plot5,plot6,plot7,plot8,ncol=2)

#Now we look at the unhappiest countries and see Rwanda is the only country that shows up in all years.  
bottom<-MeanHappy%>%arrange(Score)%>%slice(1:10)
bottom_2015<-Happiness%>%filter(Year==2015)%>%arrange(Happiness.Score)%>%slice(1:10)
bottom_2016<-Happiness%>%filter(Year==2016)%>%arrange(Happiness.Score)%>%slice(1:10)
bottom_2017<-Happiness%>%filter(Year==2017)%>%arrange(Happiness.Score)%>%slice(1:10)
bottom_2018<-Happiness%>%filter(Year==2018)%>%arrange(Happiness.Score)%>%slice(1:10)
bottom_2019<-Happiness%>%filter(Year==2019)%>%arrange(Happiness.Score)%>%slice(1:10)
bottom_2020<-Happiness%>%filter(Year==2020)%>%arrange(Happiness.Score)%>%slice(1:10)
bottom_2021<-Happiness%>%filter(Year==2021)%>%arrange(Happiness.Score)%>%slice(1:10)
bottom_2022<-Happiness%>%filter(Year==2022)%>%arrange(Happiness.Score)%>%slice(1:10)
bottom_allyears<-rbind(bottom_2015,bottom_2016,bottom_2017,bottom_2018,bottom_2019,bottom_2020,bottom_2021,bottom_2022)
table(bottom_allyears$Country)

#Now we will look at scores by top countries by year, and Rwanda score by year.
p <- ggplot(data=top_8_all_years, aes(x = Year , y = Happiness.Score)) +
  geom_line() +facet_grid(Country~.)
p

#Rwanda by year
p <-Happiness%>%filter(Country=='Rwanda')%>%ggplot(aes(x = Year , y = Happiness.Score)) + geom_line()
p

########################################################################
#CORRELATION BETWEEN FACTORS AND HAPPINESS SCORES ON ALL COUNTRIES
#We seek any relationship between happiness and various other dimensions 
########################################################################

#relationship between happiness and GDP
plot1<-Happiness%>%ggplot(aes(Happiness.Rank, Economy..GDP.per.Capita.)) + 
  geom_point(alpha = 0.5)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+labs(x=NULL)

#relationship between happiness and family 
plot2<-Happiness%>%ggplot(aes(Happiness.Rank, Family)) + 
  geom_point(alpha = 0.5)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+labs(x=NULL)

#relationship between happiness and social support 
plot3<-Happiness%>%ggplot(aes(Happiness.Rank, Social.support)) + 
  geom_point(alpha = 0.5)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+labs(x=NULL)

# relationship between happiness and life expectancy
plot4<-Happiness%>%ggplot(aes(Happiness.Rank, Health..Life.Expectancy.)) + 
  geom_point(alpha = 0.5)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+labs(x=NULL)

#relationship between happiness and freedom 
plot5<-Happiness%>%ggplot(aes(Happiness.Rank, Freedom)) + 
  geom_point(alpha = 0.5)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+labs(x=NULL)

# relationship between happiness and Governement corruption
plot6<-Happiness%>%ggplot(aes(Happiness.Rank, Trust..Government.Corruption.)) + 
  geom_point(alpha = 0.5)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+labs(x=NULL)

#relationship between happiness and generosity
plot7<-Happiness%>%ggplot(aes(Happiness.Rank, Generosity)) + 
  geom_point(alpha = 0.5)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+labs(x=NULL)

#relationship between happiness and Dystopia
plot8<-Happiness%>%ggplot(aes(Happiness.Rank, Dystopia.Residual)) + 
  geom_point(alpha = 0.5)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+labs(x=NULL)

#relationship between happiness and perception of corruption
plot9<-Happiness%>%ggplot(aes(Happiness.Rank, Perceptions.of.corruption)) + 
  geom_point(alpha = 0.5)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+labs(x=NULL)
  
#happiness & GdpHealth
plot10<-Happiness%>%ggplot(aes(Happiness.Rank, GdpHealth)) + 
  geom_point(alpha = 0.5)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+labs(x=NULL)

#happiness & GdpHealthFree
plot11<-Happiness%>%ggplot(aes(Happiness.Rank, GdpHealthFree)) + 
  geom_point(alpha = 0.5)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+labs(x=NULL)

#highest correlations
grid.arrange(plot1,plot2,plot3,plot4,plot5,plot6,plot7,plot8,plot9,plot10, plot11,ncol=3)
  

#We calculate correlation on top 5 observable correlations
corEcon<-round(cor(Happiness$Happiness.Score,Happiness$Economy..GDP.per.Capita.),digits=2)
corLifeEx<-round(cor(Happiness$Happiness.Score,Happiness$Health..Life.Expectancy.),digits=2)
corFree<-round(cor(Happiness$Happiness.Score,Happiness$Freedom),digits=2)
corGdpHealth<-round(cor(Happiness$Happiness.Score,Happiness$GdpHealth),digits=2)
corGdpHealthFree<-round(cor(Happiness$Happiness.Score,Happiness$GdpHealthFree),digits=2)
corEcon
corLifeEx
corFree
corGdpHealth
corGdpHealthFree

corGdpHealth2<-paste(c("Corr. Factor: ",corGdpHealth),collapse=" ")
corGdpHealthFree2<-paste(c("Corr. Factor: ",corGdpHealthFree),collapse=" ")

#Plot if top 2 correlation values
#relationship between happiness and GdpHealth 
plot1<-Happiness%>%ggplot(aes(Happiness.Score, GdpHealth)) + 
  geom_point(alpha = 0.5)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+labs(x=corGdpHealth2)
#Health and GdpHealthFree
plot2<-Happiness%>%ggplot(aes(Happiness.Score, GdpHealthFree)) + 
  geom_point(alpha = 0.5)+theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+labs(x=corGdpHealthFree2)

grid.arrange(plot1,plot2,nrow=2)

# For happiest countries, correlations are small.  The largest is a negative correlation with Generosity
corTopEcon<-round(cor(top_8_all_years$Happiness.Score,top_8_all_years$Economy..GDP.per.Capita.),digits=2)
corTopLifeEx<-round(cor(top_8_all_years$Happiness.Score,top_8_all_years$Health..Life.Expectancy.),digits=2)
corTopFreedom<-round(cor(top_8_all_years$Happiness.Score,top_8_all_years$Freedom),digits=2)
corTopGenerosity<-round(cor(top_8_all_years$Happiness.Score,top_8_all_years$Generosity),digits=2)
corTopGdpHealth<-round(cor(top_8_all_years$Happiness.Score,top_8_all_years$GdpHealth),digits=2)
corTopGdpHealthFree<-round(cor(top_8_all_years$Happiness.Score,top_8_all_years$GdpHealthFree),digits=2)
corTopEcon
corTopLifeEx
corTopFreedom
corTopGenerosity
corTopGdpHealth
corTopGdpHealthFree

# For unhappiest countries, correlations are small.  The largest is a negative correlation with Generosity
corBottomEcon<-round(cor(top_8_all_years$Happiness.Score,top_8_all_years$Economy..GDP.per.Capita.),digits=2)
corBottomLifeEx<-round(cor(top_8_all_years$Happiness.Score,top_8_all_years$Health..Life.Expectancy.),digits=2)
corBottomFreedom<-round(cor(top_8_all_years$Happiness.Score,top_8_all_years$Freedom),digits=2)
corBottomGenerosity<-round(cor(top_8_all_years$Happiness.Score,top_8_all_years$Generosity),digits=2)
corBottomGdpHealth<-round(cor(top_8_all_years$Happiness.Score,top_8_all_years$GdpHealth),digits=2)
corBottomGdpHealthFree<-round(cor(top_8_all_years$Happiness.Score,top_8_all_years$GdpHealthFree),digits=2)
corBottomEcon
corBottomLifeEx
corBottomFreedom
corBottomGenerosity
corBottomGdpHealth
corBottomGdpHealthFree


########################################################################
#PREDICTING HAPPINESS SCORE 
#We seek various ways of predicting happiness scores
########################################################################


#We create train set to contain years 2015 - 2021, and test_set to contain year 2022
train_set <- Happiness %>% filter(Year!='2022')%>%select(Year,Country,CtryYear, Happiness.Score,Economy..GDP.per.Capita.,Health..Life.Expectancy.)
max(train_set$Year)
test_set <- Happiness %>% filter(Year=='2022')%>%select(Year,Country,CtryYear,Happiness.Score,Economy..GDP.per.Capita.,Health..Life.Expectancy.)
min(test_set$Year)

# First model, we use the mean of train_set on test_set to get a squared loss of 1.19
HappyTrain<-mean(train_set$Happiness.Score)
HappyTrain
Sq_Loss<-mean((HappyTrain - test_set$Happiness.Score)^2)
Sq_Loss
#create table to store results
Results<-data_frame(Method="Mean Happiness Score",SqLoss=Sq_Loss)



#Data is approximately bivariate normal. 
Happiness %>% mutate(z_happy = round((Happiness.Score - mean(Happiness.Score)) / sd(Happiness.Score))) %>%ggplot()+ stat_qq(aes(sample = Happiness$Economy..GDP.per.Capita.)) +facet_wrap( ~ z_happy)

Happiness %>% mutate(z_happy = round((Happiness.Score - mean(Happiness.Score)) / sd(Happiness.Score))) %>%ggplot() +  stat_qq(aes(sample = Happiness$Health..Life.Expectancy.)) + facet_wrap( ~ z_happy)



#MODEL 2 uses the regression line based on GDP and Health to predict happiness score on the test set - Squared loss = 0.74
line_fit<-lm(Happiness.Score~Economy..GDP.per.Capita.+ Health..Life.Expectancy.,data=test_set)
line_fit

prediction <- line_fit$coef[1] + line_fit$coef[2]*test_set$Happiness.Score
Sq_Loss<-mean((prediction - test_set$Happiness.Score)^2)
Results <- Results%>% add_row(Method="Correlation Line",SqLoss=Sq_Loss)

#Plotting the regression line against test set, year 2022.
Happiness %>% 
  filter(Year %in% 2022) %>% 
  mutate(R_hat = predict(line_fit, newdata = .)) %>%
  ggplot(aes(R_hat, Happiness.Score, label = Country)) + 
  geom_point() +
  geom_text(nudge_x=0.1, cex = 2) + 
  geom_abline()

#MODEL 3 - Using the Mean Squared errors method Naive model

#define the mean squared error function
RMSE <- function(true_score, predicted_score){
  sqrt(mean((true_score - predicted_score)^2))
}

#Mean score of all countries, 2015-2021
mu_hat <- mean(train_set$Happiness.Score)
mu_hat
naive_rmse <- RMSE(test_set$Happiness.Score, mu_hat)
naive_rmse
Sq_Loss<-mean((mu_hat - test_set$Happiness.Score)^2)
Sq_Loss
Results$RMSE<-NA
Results <- Results%>% add_row(Method="Naive RMSE - RMSE using Mean",SqLoss=Sq_Loss,RMSE=naive_rmse)
Results

#MODEL 4 - Using the Mean Squared errors adding Country bias
mu <- mean(train_set$Happiness.Score) 
mu
country_avgs <- train_set %>% 
  group_by(Country) %>% 
  summarize(b_ctry = mean(Happiness.Score - mu))

predicted_ratings <- test_set %>%
 left_join(country_avgs, by='Country') %>%
  mutate( pred = mu + b_ctry)

nas<-subset(predicted_ratings,is.na(b_ctry))
nas

#remove nas from test and train sets
test_set <- subset(test_set, CtryYear!='2022_Czechia')
test_set <-subset(test_set, CtryYear!='2022_Congo')
test_set <-subset(test_set, CtryYear!='2022_Eswatini. Kingdom of')

train_set <- subset(train_set, CtryYear!='2022_Czechia')
train_set <-subset(train_set, CtryYear!='2022_Congo')
train_set <-subset(train_set, CtryYear!='2022_Eswatini. Kingdom of')

# recalculate with only countries which are in both
mu <- mean(train_set$Happiness.Score) 
mu
country_avgs <- train_set %>% 
  group_by(Country) %>% 
  summarize(b_ctry = mean(Happiness.Score - mu))



predicted_ratings <- test_set %>%
  left_join(country_avgs, by='Country') %>%
  mutate(pred = mu + b_ctry) %>%
  pull(pred)




model_1_rmse <- RMSE(test_set$Happiness.Score,predicted_ratings) 
model_1_rmse
Sq_Loss<-mean((mu - test_set$Happiness.Score)^2)
Sq_Loss

Results <- Results%>% add_row(Method="RMSE - with Country Bias",SqLoss=Sq_Loss,RMSE=model_1_rmse)

#It looks as if using the correlation line minimizes the squared loss and RMSE - with Country Bias
#minimizes the root mean squared error.
Results











```

